---
title: "Analysis.md"
author: "Cam Lu, Festus Yeboah, Chris Havenstein"
date: "August 20, 2017"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview to Analysis.md

This file, Analysis.md, includes the analysis of two different questions. Each question has included sub-questions. The questions came from the "casestudy2.pdf" provided by Dr. Jongwook Woo. Last, "Analysis.md" contains a conclusion.

* **Question 2: Orange data analysis**
    + Step **(a)**, calculate the mean and median trunk circumferences for different size of trees.
    + Step **(b)**, make a scatter plot of the trunk ciercumferences against the age of the tree. Use different plotting symbols.
    + Step **(c)**, Display the trunk circumferences on a comparative boxplot against tree. Be sure to order the boxplots in the increasing order of maximum diameter.
    
* **Question 3: Country, US temperature, and city temperature data analysis**
    + First, **(i)** find the difference between the maximum and the minimum monthly average temperatures for each country and report/visualize the top 20 countries with the maximum differences for the period since 1900.
    + Second, **(ii)** select a subset of data called “UStemp” with US land temperatures from 01/01/1990 in the Temp dataset. Use the UStemp dataset to answer the following: **(a)** Create a new column to display the monthly average land temperatures in Fahrenheit (°F). **(b)** Calculate average land temperature by year and plot it. The original file has the average land temperature by month. **(c)** Calculate the one year difference of average land temperature by year and provide the maximum difference (value) with corresponding two years.
    + Third, **(iii)** download “CityTemp” data set at box.com. Find the difference between the maximum and the minimum temperatures for each major city and report/visualize top 20 cities with maximum differences for the period since 1900.
    + Last, **(iv)** compare the two graphs in **(i)** and **(iii)** and comment it.

* **Conclusion**: 
    + Our closing thoughts are included in this section.

# Question 2. Orange data analysis

### Gathering and Cleaning the Data
First, we began by sourcing the two files that gather and clean the data for analysis. These files are **Gather.R** and **TidyData.R.**

```{r source}
source("Gather.R")
source("TidyData.R")
```

### Orange Data Variable Descriptions

* These descriptions came from the original "casestudy2.pdf" document.

Variable      |                     Description
--------------|----------------------------------------------------
  Tree        |an ordered factor indicating the tree on which the measurement is made. The ordering is according to                 increasing maximum diameter. 
  Age         |a numeric vector giving the age of the tree (days since 1968/12/31)
circumference |a numeric vector of trunk circumferences (mm). This is probably “circumference at breast height”, a                  standard measurement in forestry


### Question 2, step (a) - Mean and Median Trunk Circumferences by Tree Groups

First, we were required to calculate the mean and median of the each tree group. We used the "doBy" package to summarize the data. The "doBy" package is loaded below and you can see a sample of the data.

```{r}
library(doBy)
head(Orange)
```


Below, we provide the mean and the median of the trunk circumferences for different size of the trees. Note the "summaryBy" function from the "doBy" package.

```{r OrangeMean}
options(width =120)
summaryBy(circumference~Tree, data = Orange, FUN = c(mean, median, max, min))
     
```

### Question 2, step (b) - Scatter Plot of Trunk Circumference Against Age by Tree Groups

The "ggplot2" package was used to create the scatter plot.

```{r OrangeScatterPlot, fig.align = "center"}
library(ggplot2)
o <- ggplot(data = Orange, aes(age, circumference))
o + geom_point(aes(color = Tree, shape = Tree), size = 2.5) + geom_smooth() + labs(x = "Age(days since 1968/12/31)", y = "Tree Trunk Circumferences (mm)", title = "Orange Tree Age vs. Trunk Circumferences") +   theme(plot.title = element_text(hjust = 0.5))
```

* The scatter plot above suggests a positive relationship between age of trees (in each group) and trunk circumfrences independent of tree factor (i.e. group) of 3, 1, 5, 2, 4. It is worth mentioning that these groups are sorted in order of increasing mean circumference. There seems to be more variation between tree groups as age increases. For example, each group of tree has approximately the same tree trunk circumference when the trees have an age of 50 days. However, maximum tree trunk circumferences vary greatly between each tree group by 1600 days of age.


### Question 2, step (c) - Boxplot of Trunk Circumferences by Tree Group


* The question requests that the boxplot has each tree group ordered in terms of increasing order of maximum diameter. The boxplot below is ordered by tree group with the same order suggested in question 2, step (b). Intuitively, trees with larger circumferences will have larger diameters, but we proved the order of tree groups is the same, below.

```{r OrangeBoxplot, fig.align = "center"}

#first we need to convert the circumferences to diameters, to add a new column called diameter

#Diameter = Circumference/pi
Orange$Diameter <- Orange$circumference/pi

#Show that the scatterplot is ordered in increasing order of diameter by tree group
DiameterSummary <- summaryBy(Diameter~Tree, data = Orange, FUN = c(mean, median, max, min))
DiameterSummary
#Tree group 3 has the lowest max diameter, followed by, 1, 5, 2, and 4.

g <- ggplot(data = Orange, aes(Tree, circumference))
g + geom_boxplot(aes(color = Tree)) + labs(x = "Ordered Tree Factor", y = "Tree Trunk Circumferences (mm)", title = "Orange Tree Factor vs. Truck Circumferences") + theme(plot.title = element_text(hjust = 0.5))
```

#Question 3: Country, US temperature, and city temperature data analysis

Before answering the first part of this analysis question, please refer to the variable descriptions for the "Temp" data set, below.

### Temp Data Variable Descriptions


Variable                        |                     Description
--------------------------------|----------------------------------------------------
Date                            |   The date of average monthly temperatures measure 
Monthly.AverageTemp             |   The monthly average temperature 
Monthly.AverageTemp.Uncertainty |   The monthly temperature variation.
Country                         |   The country of monthly temperature is taken



### Question 3, part (i) - Top 20 Differences Between Maximum and Minimum Monthly Average Temperatures, by Country.

* To analyze these differences in monthly average temperatures by countries, Cam Lu created a function to find the difference between the maximum and minimum values in the "Temp" data set. When combining this function with the "summaryBy" function from the "doBy" package, this allowed us to quickly find the maximum and minimum monthly average temperature differences by country. 

* After finding these differences by country, we visualized the top 20 differences by country by using the "ggplot2" package to create a bar chart.

```{r TemperaturDiff,  warning = FALSE, fig.align = "center"}
head(temp)
myfun <- function(x){c((max(x, na.rm = TRUE)-min(x, na.rm = TRUE)), max(x, na.rm = TRUE), min(x, na.rm = TRUE))}
tempDiff <- summaryBy(Monthly.AverageTemp~Country, data = temp1900, FUN = myfun, fun.names = c("MaxMinDifference", "Max", "Min"))
tempDiffTop20 <- head(tempDiff[order(tempDiff$Monthly.AverageTemp.MaxMinDifference, decreasing = TRUE),], 20)
#update the index on the first column
rownames(tempDiffTop20) <- 1:nrow(tempDiffTop20)
tempDiffTop20

#create the bar chart in ggplot2
g  <- ggplot(tempDiffTop20, aes(x = reorder(Country, -Monthly.AverageTemp.MaxMinDifference), y= tempDiffTop20$Monthly.AverageTemp.MaxMinDifference))
g1 <- g + geom_bar(stat = "identity", fill="blue") + theme(axis.text.x = element_text(size = 7.5, angle = 55, hjust = 1)) +  xlab("Country") + ylab("Monthly Avg. Temperature Difference (F)") + ggtitle("Top 20 Max. and Min. Avg. Temperature Differences By Countries")
g1

```


### Question 3, part (ii) - "USTemp" Subset From "Temp" Data Set, since 1990

The "Temp" data set was used to create a subset of the country temperature data, where the country was "United States" and only containing values since 1990.

This question has three sub-questions.

#### Question 3, part (ii), sub-question (a) - Create Monthly Average Temperature Column in Fahrenheit

* To quickly and accurately convert the "Monthly.AverageTemp" column to fahrenheit (from celsuis), we used the "weathermetrics" package. 

* The first 6 rows containing the new column "Monthly.AverageTemp.Fahrenheit" are provided below. Please note the new column was added as the last column in the dataset (possibly requiring you to scroll to the right).

```{r Fahrenheit, fig.align = "center"}
##Enable the weathermetrics package to convert temperature from Celsius to Fahrenheit
##Need to run install.packages(“weathermetrics") to install package first if have not.

library(weathermetrics)

##Create a new column to display the monthly average land temperatures in Fahrenheit (°F). 

UStemp$Monthly.AverageTemp.Fahrenheit <- celsius.to.fahrenheit(UStemp$Monthly.AverageTemp)
head(UStemp)
```

#### Question 3, part (ii), sub-question (b) - Time Series plot of US Average Land Temperatures by Year

* To provide the time series plot of the US average land temperatures by year, we had to aggregated the  monthly data by year. The "aggregate" function was used below to do this. The yearly aggregated values were stored in the dataframe "YearlyAverageTemp".

* Then, the plot was created by using "ggplot2".

```{r UStimeseriesPlot, fig.align= "center"}
##Calculate average temperature for each year.
options(digits=5)
YearlyAverageTemp <- aggregate(UStemp[, 6], list(UStemp$Year), mean, na.rm = TRUE)
names(YearlyAverageTemp) <- c("Year", "Average.Temp")
YearlyAverageTemp
p <- ggplot(data = YearlyAverageTemp, aes(Year, Average.Temp, group = 1))
p + geom_point() + geom_line(color = 'red') + labs(y = "Yearly Avg. Temperature In Fahrenheit (F)", 
                                                   title = "US Land Temperature From 1990") +
                                        theme(axis.text.x = element_text(size = 7.5, angle = 55, hjust = 1))
```


#### Question 3, part (ii), sub-question (c) - One Year Differences of US Average Land Temperatures

* To create the one year differences of US average land temperatures, we added a new column to our YearlyAverageTemp dataframe called "DiffTempFromPreviousYear".

* The results are displayed, below. Please note that the year "1990" has a "NA" value. This is because there wasn't a previous year in the "USTemp" subset before 1990.

```{r USyearDiffs, fig.align= "center"}
YearlyDiffTemp<-round(diff(YearlyAverageTemp$Average.Temp), 3)
YearlyAverageTemp$DiffTempFromPreviousYear <- c("NA", YearlyDiffTemp)
YearlyAverageTemp

```

### Question 3, part (iii) - "CityTemp" Top 20 Differences Between Maximum and Minimum Monthly Average Temperatures, by Major Cities

* To analyze these differences in monthly average temperatures by major cities, we reused Cam Lu's function to find the difference between the maximum and minimum values in the "CityTemp" data set. When combining this function with the "summaryBy" function from the "doBy" package, this allowed us to quickly find the maximum and minimum monthly average temperature differences by major cities. 

* After finding these differences by country, we visualized the top 20 differences by country by using the "ggplot2" package to create a bar chart.

```{r CityTemperaturPlot,fig.align = "center"}

head(CityTemp1900)
CityTemp1900MaxMinDiff <-  summaryBy(Monthly.AverageTemp~City, data = CityTemp1900, FUN = myfun)
CityTemp1900MaxMinDiffTop20 <- head(CityTemp1900MaxMinDiff[order(CityTemp1900MaxMinDiff$Monthly.AverageTemp.FUN1, decreasing = TRUE),],20)
names(CityTemp1900MaxMinDiffTop20) <- c("City", "MaxDiffTemp", "MaxTemp", "MinTemp")

#update the index on the first column
rownames(CityTemp1900MaxMinDiffTop20) <- 1:nrow(CityTemp1900MaxMinDiffTop20)
CityTemp1900MaxMinDiffTop20

#creating the bar chart in ggplot2 
f  <- ggplot(CityTemp1900MaxMinDiffTop20, aes(x = reorder(City, -MaxDiffTemp), y= MaxDiffTemp))
f1 <- f + geom_bar(stat = "identity", fill="red") + theme(axis.text.x = element_text(size = 7.5, angle = 55, hjust = 1)) + xlab("City") + ylab("Monthly Avg. Temperature Difference (F)") + ggtitle("Top 20 Max. & Min. Avg. Temperature Differences By Cities")
f1

```

### Question 3, part (iv) - "Temp" and "CityTemp" Plot Comparison, Countries Vs. Major Cities

* By using the bar charts we have already created, we created a side-by-side comparison of the countries and cities bar charts.

* To do this, we applied some slight formating changes to previous "ggplot2" plots and used the package "cowplot" to easily create the side-by-side comparison.

* Our comments on these two plots are provided below the comparison.


```{r ComparePlots, message=FALSE ,fig.align= "center"}

#Add additional formatting for side-by-side plots

g1 <- g1+theme(axis.text=element_text(size=5),
              axis.title=element_text(size=8,face="bold"),
              plot.title = element_text(size=8))

f1 <- f1+theme(axis.text=element_text(size=5),
               axis.title=element_text(size=8,face="bold"),
               plot.title = element_text(size=8))

#install.packages("cowplot")
library("cowplot")

#display the plots for comparison
plot_grid(g1,f1,
          ncol = 2, nrow = 1)

```

* Our comments for the side-by-side comparison:
    + The countries on the left bar chart often have representative major cities on the right city bar chart.
    + There are many asian cities on the right bar chart. 
    + On the left bar chart for countries, 
    + Many of the countries listed have locations with extremely code climates within their country.
    + The countries and cities are not in the same order. Intuitively, this makes sense because a city may not represent the country that it is a member of.
    + The range of the average monthly temperature differences are very similar. For countries, the range is about 0 to 48 degrees fahrenheit. For Cities, the range is about 0 to 47 degrees fahrenheit.


# Conclusion

* In this conclusion we summarize the findings from questions 2 and 3. 

* In question 2, first, we calculated the mean and median tunk circumferences. We found that tree group "3" had the smallest mean and median trunk circumferences and tree group "4" had the largest mean and median trunk circumferences. Second, we created a scatter plot of the trunk circumferences, with difference symbols for each tree group and age pair (with a total n=35). Initially, each of these tree groups had very similar tree circumferences, but as their age increased, the variation between tree groups increased. Last, we created a boxplot of the tree circumferences against each tree group. The order of maximum diameter of each tree was the same as the previous order of tree circumferences (in increasing order: 3,1,5,2,4) and this was how the boxplots were ordered.

* In question 3, first, we found the maximum and minimum differences in monthly average temperatures by country and provided a bar chart for the top 20 countries. Many of these were Eastern European or Asian countries (14/20). Second, we examined the "UStemp" subset of the "Temp" country data set. We plotted the yearly US average temeperatures from 1990 to 2012. There appears to be a slight increasing trend in US average yearly temperature, although the range of the plot is narrow (48-53 degrees fahrenheit). Third, we created a bar chart of the top 20 major cities maximum and minimum differences in monthly average temperatures. The highest differences were in major cities that had very low minimum temperatures. Then last, we compared these same differences between the top 20 countries and major cities. The highest and lowest temperature differences in our two bar charts were about the same.
    
